Dark matter halo shape at $z=0$ in the Auriga simulations: radial evolution and alignment with the stellar disk

### 2019-07-19 Document/paper.tex
edit

\citep{Debattista08} reported a similar twisting effect, but only at radii smaller
than the disk radius. 
At radii larger than the disk radius, they find, the dark matter
shells stop twisting. 
\citep{JingSuto02} reported shell alignment twisting in their dark
matter only simulations as measured in the direction changes of large
and medium shape axis. 
They only had three high resolution MW-like halos (with
$\sim 10^{6}$ inside the virial radius) and could not make a
statistical statement about the impact of this effect.
However, they the twisting as an artifact resulting from high values of the
$c/b$ ratios that make the determination of the medium axis boisy.
Could then high triaxility values explain the twisting as a result of
erratic axis directions? 
If the triaxiality were the main culprit behind the twisting, the
correlation between the two quantities should be constant at every
radii.
In Figure \ref{fig:alignment_correlations} we already showed that this
correlation strongly depends on radius, we conclude that high
triaxiality values cannot be the full explanation.
Furthermore the twisting happens in such a way that it allows and
almost perfect alignment between the stellar disk at $0.1R{200}$ and
the halo shape measured at $0.25R_{200}$

The measurements we have done in the simulations can be compared
against observational results.
We use the observational constraints by \cite{LM10} and \cite{Bovy16}
to place our results in an observational context.
\cite{LM10} used observations of the long Sagittarius tidal stream to
constrain the shape of the gravitational potential.
Their point of depart is that previous studies that assumed an
axisymetric Galactic potential were not able to fit all the available
dynamic constraints for the Sagittarius stream, therefore they
proposed a rigid triaxial potential, with coaxial potential
ellipsoids, for the dark matter component.  
Their results are able to constrain the triaxility of this potential
component, translating these results into a triaxiality of the density
contours they report $(c/a)=0.44$ and $(b/a)=0.97$ at a radius of
$\sim 40$kpc. They do not report any uncertainties for these two
values. 
Looking at their plots of their quality of fit criterion as a function
of dark halo axial scales (Figure 5), we choose a conservative $10\%$
relative uncertainty, close to the uncertainty in those parameters.
The element that made the results by \cite{LM10} (LM10 hereafter) surprising is that
the minor axis of the halo shape is parallel, not perpendicular, to
stellar disk plane.

The results by \cite{Bovy16} (BBFK16 hereafter) have the same general approach but use
instead the GD-1 \citep{2006ApJ...641L..37G} and Pal 5 \citep{2009AJ....137.3378O}
streams to constraint the shape of the dark matter component of the
galacic halo potential.
They use general models with large degrees of freedom for the galactic
potential in order to measure to what extent these two streams are sensitive
to the triaxiality of the dark matter halo component.
The DM component is written directly as a triaxial density profile
with coaxial ellipsoids and the corresponding potential is found by
numerical integration.
They find that the width of the Pal 5 stream constraints $b/a\approx
1$ and therefore fix it to $b/a=1$, finally they report their most stringent
constrain of $c/a=0.93\pm0.16$ at a radius of $\approx 19$kpc from the
galactic center.

Our Figure \label{fig:observations} shows an explicit comparison in
the $c/a$-$b/a$ plane against the results by LM10 and BBFK16. 
We find $6$ halos with $b/a<0.93$ and $c/a>0.77$ that could be
considered consistent with their shape constraints 
by BBFK16.
In contrast, none of the simulated halos is consistent with the LM10
results.

The results by LM10 would place then the dark matter halo of our Milky
Way as an extreme outlier in the $\Lambda$ CDM model. 
This extreme prolateness also correlates with the extreme
triaxiality of the 11 classical satellites of the MW ($c/a\approx
0.2$, and $b/a\approx0.9$) with an spatial distribution 
also oriented perpendicular to the MW plane, another highly unusual
feature in the $\Lambda$ CDM model \citep{2018MNRAS.478.5533F}

>>>>

\cite{Debattista08} reported a similar twisting effect, but only at radii smaller
than the disk radius. 
At radii larger than the disk radius, they find, the dark matter
shells stop twisting. 
\cite{JingSuto02} also reported shell alignment twisting in their dark
matter only simulations as measured in the direction changes of large
and medium shape axis. 
They only had three high resolution MW-like halos (with
$\sim 10^{6}$ inside the virial radius) and could not make a
statistical statement about the impact of this effect.
However, they the twisting as an artifact resulting from high values of the
$c/b$ ratios that make the determination of the medium axis boisy.
Could then high triaxility values explain the twisting as a result of
erratic axis directions? 
If the triaxiality were the main culprit behind the twisting, the
correlation between the two quantities should be constant at every
radii.
In Figure \ref{fig:alignment_correlations} we already showed that this
correlation strongly depends on radius, we conclude that high
triaxiality values cannot be the full explanation.
Furthermore the twisting happens in such a way that it allows and
almost perfect alignment between the stellar disk at $0.1R{200}$ and
the halo shape measured at $0.25R_{200}$

We now use the results reported by \cite{LM10} and \cite{Bovy16}
to place our results in an observational context.
\cite{LM10} used observations of the Sagittarius tidal stream to
constrain the shape of the gravitational potential.
Their point of depart is that previous studies that assumed an
axisymetric galactic potential were not able to fit all the available
dynamic constraints for the Sagittarius stream, therefore making
necessary the use of a rigid triaxial potential with coaxial potential
ellipsoids for the dark matter component.  
Their results constrain the triaxility of this potential
component. 
They also translate ther results into a triaxiality of the density
contours (that could be compared against our results)
 to be $(c/a)=0.44$ and $(b/a)=0.97$ at a radius of $\sim 40$kpc. 
They do not report any uncertainties for these two values. 
Looking at their plots for the quality of fit criterion as a function
of dark halo axial scales (their Figure 5), we choose a conservative $10\%$
relative uncertaintyy
One surprising element in their results is that  the major axis of the
halo shape is perpendicular to the stellar disk plane.

The results by \cite{Bovy16} have the same general approach but use
instead the GD-1 \citep{2006ApJ...641L..37G} and Pal 5 \citep{2009AJ....137.3378O}
streams to constraint the shape of the dark matter component of the
galacic halo potential.
They use general models with many degrees of freedom for the galactic
potential in order to measure to what extent these two streams are sensitive
to the triaxiality of the dark matter halo component.
The DM component is written directly as a triaxial density profile
with coaxial ellipsoids and the corresponding potential is found by
numerical integration.
They find that the width of the Pal 5 stream constraints $b/a\approx
1$ and therefore fix it to be $b/a=1$ exactly.
Using that value they report their most stringent constrain of
$c/a=0.93\pm0.16$ at a radius of $\approx 19$kpc from the galactic
center.

Our Figure \label{fig:observations} shows an explicit comparison in
the $c/a$-$b/a$ plane of our results in MHD simulations against the
results by LM10 and BBFK16.  
We find six MHD halos with $b/a<0.93$ and $c/a>0.77$ that could be
considered consistent with their shape constraints  by BBFK16, while
only one outlier DMO halo is consistent with those constraints.
In contrast, none of the simulated halos (MHD nor DMO) is consistent
with the LM10 results. 
The change of triaxility with radius in our simulations cannot account
for these two extremely different observational constraints at
different radii.

The results by LM10 could then place the dark matter halo of our Milky
Way as an extreme outlier in the $\Lambda$ CDM model. 
This extreme prolateness also correlates with the extreme
triaxiality of the 11 classical satellites of the MW ($c/a\approx
0.2$, and $b/a\approx0.9$) with an spatial distribution 
also oriented perpendicular to the MW plane, another highly unusual
feature in the $\Lambda$ CDM model \citep{2018MNRAS.478.5533F}.
